---
title: "Method2_Logit_Interactions"
author: "Clayton Fogler"
format: html
editor: visual
---

In our first method, we used logistic regression to find our best-fit model for predicting success based on RP, SITUATION, DN, DIST, and PERSONNEL. We found that a logistic Regression Model using DN (Down), DIST (Distance to go), and RP (whether the play was a run or pass) was our best model based on AIC (resulted in an AIC of 1095.1). This model (called logBEST), resulted in a classification rate of 0.632, meaning our model correctly predicted 63.2% of the plays.

Our next method is going to be introducing interaction terms to our models to see if they can yield a higher classification rate. 

Just like we did with the original logistic regression models, we will build a model with all our terms that include pairwise interactions for all 5 variables:

```{r}
logALL_inter <- glm(SUCCESSFUL ~ (RP + SITUATION + DIST + DN + PERSONNEL)^2, data = all_seasons_logit, family = binomial)
summary(logALL_inter)
```

Just like we did with the original logistic regression models, we will use the stepwise function to sort through all of the combinations of our 5 variables to see which interaction model would be the best:


```{r}
logBEST_inter <- step(logALL_inter, direction = "both", trace = FALSE)
summary(logBEST_inter)
```

*It seems that the best possible model using our 5 variables to predict success, based on AIC and using interaction terms, is a model with RP, DIST,DN, RP:DIST interaction, and DIST:DN interaction (AIC = 1090.6)*


*It seems that the best possible model using our 5 variables to predict success, based on AIC, is a model with RP, DIST, and DN (AIC of 1095.1).*

The AIC of our model including interaction (1090.6) was better than the model without interaction (1095.1), but lets see if its classification rate was also better. 

```{r}

# predicted probabilities for Success: 
pred_probs_inter <- predict(logBEST_inter, type = "response")

# Converting the predicted probabilities to either successful or not successful 
# (turning a predicted probability of 0.51 to a 1 for successful, or a 0.49 to 0 for not successful)
pred_class_inter <- ifelse(pred_probs_inter > 0.5, 1, 0)

# Now we grab our actual values to compare to the predicted values we just found
#already ran code during Method1, but just wanted to add it here to help show the process
actual <- all_seasons_logit$SUCCESSFUL

# classification rate calculation:
classification_rate_inter <- mean(pred_class_inter == actual)
classification_rate_inter

```

When we added an interaction between RP and DIST, and an interaction between DIST and DN, to our best logistic regression model from method 1, we find a classification rate of 0.638, which means our model correctly predicted 63.8% of the plays. This is just .6% better than the model without interaction. What this means is overall, adding in interactions had very little effect to our logistic regression model. 


For fun and visualization, we can graph our interaction predicted values based off our logistic regression model:

```{r}
regression_plot <- all_seasons_logit |>
  mutate(
    pred_probs_inter = pred_probs_inter
  )
```

```{r}
graph_interaction <- ggplot(regression_plot, aes(x = DIST, y = pred_probs_inter, color = RP)) +
  geom_point() + # each predicted observation
  geom_smooth() + # smooth line
  labs(y = "Predicted Probability of Success", x = "Down", color = "Run/Pass") +
  theme_minimal()

graph_logit + graph_interaction
```

